---
title: "Introduction to Bayesian Statistics"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: css/learnr-theme.css
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
if ("learnr" %in% (.packages()))
  detach(package:learnr, unload = TRUE)
library(learnr)
knitr::opts_chunk$set(echo = FALSE)

## Save package names as a vector of strings
pkgs <-  c()

## Install uninstalled packages
lapply(pkgs[!(pkgs %in% installed.packages())], 
       install.packages,
       repos='http://cran.us.r-project.org')

## Load all packages to library and adjust options
lapply(pkgs, library, character.only = TRUE)
```

## Introductions

### Hi, I'm Denis

- Senior Research Fellow, Data and Methods Unit, MZES
- Dr. phil. (political science), Humboldt University of Berlin (2018)
- I work on:
    - political consequences of spatial inequalities and residential mobility
    - party competition
    - voting behavior
    - quantitative social science (both pedagogical and methodological)

### Hi, I'm Franziska
 
- Master Student in Political Science, University of Mannheim
- B.A. (Political Science and Sociology), Ruprecht Karls Universität Heidelberg (2023)
- Student Assistant at ALLBUS, GESIS Leibniz Institute for the Social Sciences, Mannheim
- I’m interested in:
    - International Relations
    - Global Governance
    - European Union

### Who are you?

- Your name
- Your discipline and current position
- What brought you here

## Introduction to Bayesian Statistics

###  Learning Objectives

1. Contrast frequentist and Bayesian inference paradigms
1. Understand when and why researchers go Bayesian
1. Understand the basics of analytical and (especially) numerical Bayesian inference
1. Understand the costs and benefits of doing applied Bayesian data analysis
1. Perform applied Bayesian data analysis using the R package `brms` (Bayesian Regression Models using 'Stan')

### The end goal

Empower you to 

- implement
- execute 
- process
- scrutinize
- understand

Bayesian data analysis using pre-implemented tools.

### Advanced Bayesian Statistical Modeling Using R and Stan

This course's older sibling will take place May 20-22, 2025, in the GESIS Training Program.

The key differences in terms of focus:

- Stronger focus on statistical modeling
- Focus on building, as opposed to using models

The key differences in terms of prerequisites and technicality:

- Stronger focus on the mathematical foundations of probability distributions
- Stronger focus on programming
- Stronger focus on the Stan language, as opposed to the `brms` front-end

### A quick note on mathematical notation

- The teaching materials for this course use mathematical notation
- It is helpful, albeit not strictly necessary that you understand it:
    - Mathematical notation is a compact and universal tool to convey concepts...
    - ...but what really matters is that you understand the intuition of the underlying concepts...
    - ...and you will not be asked to translate formulas into code.
- If anything is unclear: Please ask at any time!

## Course materials

### GitHub

All course materials are available on my GitHub: https://github.com/denis-cohen/bayesintro 

### Interactive tutorials

The primary course materials come as `learnr` tutorials wrapped in an R package, `bayesintro`.

You can use the interactive materials to:

- click along during the lectures
- work on your lab exercises

To download, install, and use the interactive materials, run this code that you can copy from GitHub and paste into R:

```{r, eval=FALSE, echo=TRUE}
# Detach if loaded
if ("bayesintro" %in% (.packages())) {
  detach(package:bayesintro, unload = TRUE)
}

# Uninstall if installed
if ("bayesintro" %in% installed.packages()) {
  remove.packages("bayesintro")
}

# Install if not installed
if (!("devtools" %in% installed.packages())) {
  install.packages("devtools")
}

# Load from GitHub
library(devtools)
devtools::install_github("denis-cohen/bayesintro")

# Load to library
library(bayesintro)

# Run tutorials (one at a time)
learnr::run_tutorial("00-00-int", package = "bayesintro")
learnr::run_tutorial("01-01-lec", package = "bayesintro")
learnr::run_tutorial("01-02-lec", package = "bayesintro")
learnr::run_tutorial("01-03-lab", package = "bayesintro")
learnr::run_tutorial("02-01-lec", package = "bayesintro")
learnr::run_tutorial("02-02-lec", package = "bayesintro")
learnr::run_tutorial("02-03-lab", package = "bayesintro")
```

### Static course materials

#### Course notes

I have also rendered the `learnr` tutorials (minus the interactive components) as static documents:

- Lecture notes come in PDF format
- Lab notes come in HTML format in two versions: with and without code-based solutions and outputs 

Along with the data used in this course, these materials are available:

1. from the GESIS Moodle course
1. from my GitHub (see subfolders `lecture-materials`, `lab-materials`, and `data`)

## Schedule

### Sessions

| Session | Session          | Topics                                                                             |
| :-----: | ---------------: | :----------------------------------------------------------------------------------|
|    1.1  | Wed, 09:00-10:30 | *Lecture:* R and Frequentist Inference: A Refresher                            |
|    1.2  |      10:45-12:15 | *Lecture:* Bayesian Fundamentals                                               |
|    1.3  |      13:45-15:15 | Lab session Day 1                                                              |
|    2.1  | Thu, 09:00-10:30 | *Lecture:*Applied Bayesian Statistics I: Basics & Workflow                     |
|    2.2  |      10:45-12:15 | *Lecture:*Applied Bayesian Statistics II: Using brms                           |
|    2.3  |      13:45-15:15 | Lab session: Exercises and solutions                                           |
|    3.1  | Fri, 09:00-11:00 | Applied Bayesian Statistics III: Implementing a small data project using brms  |    
|    3.2  |      11:15-12:15 | Wrap-up                                                                        |
|         |                  | - Review of challenges and potentials encountered in working on your data projects |
|         |                  | - Outlook: Moving forward as Bayesian practitioners                                |
|         |                  | - Course evaluations and feedback                                                  |

### Structure of our sessions

1. Lecture:
    - Input
    - Dedicated Q&A (but feel free to ask for clarification at any point!)
2. Lab: Hands-on exercises (in small teams)
    - You will be assigned to small teams in breakout rooms
    - I will be in the main meeting, available for questions
    - We will jointly discuss your answers and the solutions in the main meeting

### Session 3.1: Implementing a small data project using brms

- Becoming an active Bayesian practitioner requires practice!
- The challenges you may encounter along the way will likely be idiosyncratic
- Therefore, Session 3.1 gives you the chance to implement the Bayesian workflow using your own data
- This can be an analysis you are about to run for the first time or a replication of your or others' existing work, as long as:
    - Pre-processing is complete: Your data is ready to go for data analysis
    - Neither your data nor your model are highly complex (if they are: start with a small subset of the data, gradually build up model complexity)
- I will be available for questions throughout Session 3.1
- We will jointly discuss your experiences and the main challenges you encountered at the start of Session 3.2

### RStan/brms installation: Any problems?

Did you successfully get RStan to run on your machine, incl. `example(stan_model, package = "rstan", run.dontrun = TRUE) `?

- If yes: Great! You're all set!
- If no: No worries, we will start using `brms` *tomorrow*. I will stick around after 15:15 today in case you need assistance.

## Questions?




